\begin{acknowledgements}\label{acknow}
%\dsp

Doing a Ph.D. eight time zones away from campus while juggling a professional career and family should have been almost impossible. But it wasn't. That transformation from the abstract to the achievable was made possible by my amazing advisors, Dr. Bahram Gharabaghi and Dr. Jesse Th\'e. Their support, encouragement and mentoring made these three years possible by arranging remote video classes to meet my course requirements, providing expert advice, recommending appropriate journals to submit to, and occasionally providing a necessary kick in the backside.

The other members of my committee, Dr. Graham Taylor and Dr. Ed McBean, gave me the programming and statistical tools I needed to complete this program, as well as a new passion in machine learning and enjoyment in Python.

I am grateful beyond measure to the Natural Sciences and Engineering Research Council of Canada (NSERC), Mitacs, and Lakes Environmental for their funding and financial support and the University of Guelph's Department of Engineering for the flexible and kind support, especially when I let registration dates slip.

While much of the data used in this thesis was collected during a United Nations Development Program project, it would be ungrateful of me not to thank my colleagues (and friends) at the Kuwait Environment Public Authority, Department of Air Quality and Follow-Up for their assistance.

For the UAS traffic data collection, my thanks go out to the Kuwait Traffic Safety Society for organizing flight approvals with the Ministry of Interior and being very enthusiastic with all my work.

The management at Integrated Environmental Solutions was overwhelmingly supportive of my program - authorizing any time off I needed and letting me use office resources (and time) to do some of my analytical work. My excellent staff assisted me in countless ways, but mostly by doing their jobs well, so I could focus on my studies.

My parents, sister, and sons also played a collective part, even while located on another continent. They provided endless moral and material support, especially during my trips back to North America. My parents especially instilled the necessary encouragement and love of learning that all great families share.

My beautiful and loving partner, Manuela, provided a fantastic environment (and great food) to complete my work in- all while putting up with my late hours, self-absorbed studying, distracted moods, and steady streams of curses when my code did not work, with patience and understanding.

Lastly, some people may complete their studies without dogs and/or cats, but I would not recommend it. My girls were always there and always ready to lift my spirits.

It's been a wonderful journey.

\end{acknowledgements}
%\dsp
